---
title       : A Canadian Language Map
subtitle    : Population by knowledge of official languages
author      : Yangang Chen
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, deckjs, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## Overview of the App

* Motivation:

Canada has two official languages: English and French. My Shiny app trys to answer the following questions: Where do Francophone Canadians live? How many Canadians are bilingual? Do people in big cities (e.g. Toronto, Vancouver) speak different languages? ......

* The objective:

Visualize and summarize the population of English/French/both/neither speakers in each city-level division

* The link:

https://yangangchen.shinyapps.io/CanadaLanguage/



---
## Basic Interface

* Part I: Interactive Map

The size of the circles represents the population size for the English/French/both/neither speakers.

The intensity of color of the circles represents the percentage of the chosen speakers among the total population.

The pie chart shows the percentage of the English/French/both/neither speaker in a chosen province/territory.

* Part II: Data Explorer

Search engine for the original data downloaded from Statistics Canada. Very straightforward!



---
## Example: Pie Chart in the Interactive Map

```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(data.table)

data <- fread("./files/CanadaLanguage_Clean.csv", stringsAsFactors = FALSE)
data[,Total:=as.numeric(Total)]
data[,TotalPercent:=as.numeric(TotalPercent)]
data[,TotalChangePercent:=as.numeric(TotalChangePercent)]
data[,ENG:=as.numeric(ENG)]
data[,ENGPercent:=as.numeric(ENGPercent)]
data[,ENGChangePercent:=as.numeric(ENGChangePercent)]
data[,FRA:=as.numeric(FRA)]
data[,FRAPercent:=as.numeric(FRAPercent)]
data[,FRAChangePercent:=as.numeric(FRAChangePercent)]
data[,ENGFRA:=as.numeric(ENGFRA)]
data[,ENGFRAPercent:=as.numeric(ENGFRAPercent)]
data[,ENGFRAChangePercent:=as.numeric(ENGFRAChangePercent)]
data[,NO:=as.numeric(NO)]
data[,NOPercent:=as.numeric(NOPercent)]
data[,NOChangePercent:=as.numeric(NOChangePercent)]

ageBy <- "1"
provinceBy <- "Canada"

subdata <- data[data$Age==ageBy & data$Province==provinceBy,
                c("Province","Division",
                  "Age","ENG","FRA","ENGFRA","NO"), 
                with=FALSE]
subdata <- data.frame(
    Language=c("English","French","Both languages","Neither language"),
    Value=c(sum(subdata$ENG), sum(subdata$FRA), sum(subdata$ENGFRA), sum(subdata$NO)) )
```
```{r}
library(ggplot2)
ggplot(data=subdata, aes(x=factor(1), y=Value, fill=Language) ) +
    geom_bar(width=1, stat="identity") +
    coord_polar(theta="y") + ylab("") + xlab("") + labs(fill="") + 
    theme(axis.ticks = element_blank(), panel.grid  = element_blank(), 
          axis.text = element_blank(), legend.position = "right") + 
    ggtitle(paste0("Statistics of ",provinceBy))
```



---
## Conclusions

* The biggest Canadian cities: Toronto, Montreal, Vancouver, Ottawa, Calgary, etc.

* Most of Canadians are anglophone (speak English).

* Most of Francophone Canadians live in the province of Quebec.

* Bilingual Canadians concentrate in Montreal, Ottawa, the border of Quebec/Ontario, and Northern New Brunswick.

* People in big cities (e.g. Toronto, Vancouver) may speak neither official languages.

* There are other interesting discoveries from the app, which will be left for you :)
